Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-02-25 paper.tex
More fixes

As an application of parabolic factorizations, we deduce several {\it subgroup factorizations} of finite width for relative Chevalley groups.
For example, as a corollary of Bass--Kolster decomposition we obtain upper estimates of the number of factors in the presentation of a classical Chevalley group
 as a product of its subsystem subgroups $\SL(2, R, I)$, see~\cref{thm:SL2width}.
As an application of~\cref{thm:DennisVaserstein} we prove the following theorem inspired by the main result of~\cite{Nik07}.
\begin{thm}\label{thm:spin-sln-prod} Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$. \end{thm}
As another application, we obtain results on the bounded generation of the relative elementary group.
Recall from~\cite[Theorem~2]{Va86} that $\E(\Phi, R, I)$ can be generated by the set of Tits--Vaserstein generators
 $z_\alpha(s, \xi) =  x_{-\alpha}(-\xi) x_\alpha(s) x_{-\alpha}(\xi)$, where $s\in I$ and $\xi\in R$
(in fact, a smaller generating set suffices, see~\cref{prop:Stepanov-theorem}).
In section~\ref{sec:boundgen} we deduce from an earlier result of O.~Tavgen that the width $\E(\Phi, R, I)$ with respect to this generating set is
 finite for Dedeking rings of arithmetic type having a real embedding.
Moreover, as a further application of parabolic factorizations, we obtain in~\cref{thm:width} concrete estimates of such width
 for ideals of the ring $\mathbb{Z}[\sfrac{1}{p}]$ and also for ideals of stable rank $1$.

@@

\textsc{Case $\Phi=\rD_\ell$.} From the proof of \cref{item:asrUnipD} it follows that there exists $h_1 \in \EP_\ell(R, I)$ such that for $v'=h_1\cdot v$ one has $v'_2=v'_1-1\in I$.
Clearly, for $v'' = z_{-\alpha_{2}}(-v'_2, 1)\cdot v'$ one has $v''_1=1$, hence by \cref{lemma:Chevalley-Matsumoto} there exists $h_2 \in \U(\Phi^-, I)$ such that the element
$g'=h_2 \cdot z_{-\alpha_{2}}(-v'_2, 1) \cdot h_1 \cdot g$ fixes $v^+$. 
Using the Levi decomposition we can write $g'=h \cdot y \cdot x \cdot g$ for some $y\in\U(\Sigma^-_\ell, I)$, $x \in \U(\Sigma^+_\ell, I)$, $h\in\E(\Delta_\ell, R, I)$.
It is clear that $x$, $y$ are the desired elements.
\end{proof}

@@

Let $G$ be a group and $A$ its subset. Denote by $L(A)$, $R(A)$ the left and the right stabilizers of $A$ and by $N(A)$ the normalizer of $A$:
\begin{align*} 
 L(A) & =  \left\{ g\in G\ \middle |\ g \cdot A = A \right\}\!,\\
 R(A) & =  \left\{ g\in G\ \middle |\ A \cdot g = A \right\}\!,\\
 N(A) & =  \left\{ g\in G\ \middle |\ g \cdot A \cdot g^{-1} \cup g^{-1} \cdot A \cdot g \subseteq A \right\}.
\end{align*}
It is easy to see that $L(A)$, $R(A)$ and $N(A)$ are subgroups of $G$. The following obvious lemma will be used several times in the sequel.
\begin{lemma} \label{lem:LN} The subgroup $L(A)$ is normalized by $N(A)$. Moreover, one has $R(A) \cap N(A) \subseteq L(A)$, $L(A) \cap N(A) \subseteq R(A)$. \end{lemma}
\begin{comment}
\begin{proof}
Let $h \in L(A)$, $g \in N(A)$. The first claim follows from the following chain of inclusions:
\begin{equation} h^g \cdot A = g^{-1} \cdot h \cdot g \cdot A \cdot g^{-1} \cdot g \subseteq g^{-1} \cdot h \cdot A \cdot g \subseteq {A}^g \subseteq A. \end{equation} 
Now if, in addition, $g \in R(A)$ we obtain
$g \cdot A = g \cdot A \cdot g^{-1} \cdot g \subseteq A \cdot g \subseteq A,$ which implies the second claim.
\end{proof}
\end{comment}

@@

Denote by $A$ the product of subgroups in the right-hand side of~\eqref{eq:gauss1}.
First of all, notice that for $h \in \Hh(\Phi, R, I)$, $\beta \in \Phi$ and $\xi \in R$ one has $x_\beta(\xi)^h = x_\beta(\xi + s\xi)$ for some $s\in I$.
From this and the assumption of the proposition we obtain for $\alpha \in \Delta_1 \cup \Delta_\ell$, $\xi \in R$ that
\begin{multline} \nonumber
 A^{x_\alpha(\xi)} = (\Hh(\Phi, R, I) \cdot A(\Phi^+, I))^{x_\alpha(\xi)} \subseteq (\Hh(\Phi, R, I) \cdot \E(\Delta_i, R, I) \cdot A(\Sigma_i^+, I))^{x_\alpha(\xi)} \subseteq \\
  \subseteq \Hh(\Phi, R, I) X_{\alpha}(I) \cdot \E(\Delta_i, R, I) \cdot A(\Sigma_i^+, I) \subseteq \Hh(\Phi, R, I) \cdot \Hh(\Delta_i, R, I) A(\Delta_i^+, I) \cdot A(\Sigma_i^+, I) \subseteq A.
\end{multline}
On the other hand, for $\beta \in \Phi^+$ we have
\begin{equation} \nonumber
 X_{\beta}(I) \cdot A \subseteq \Hh(\Phi, R, I) \cdot X_{\beta}(I) A(\Phi^+, I) \subseteq A.
\end{equation}
Thus, $A$ is normalized by root subgroups $X_\alpha(R)$, $\alpha \in \Delta_1 \cup \Delta_\ell$ and therefore is normalized by $\E(\Phi, R)$.
On the other hand, from $\U(\Phi^+, I) \cdot A \subseteq A$ and~\cref{lem:LN} we conclude that $\mathcal{X} \cdot A \subseteq A$ and hence
 that $\E(\Phi, R, I) \cdot A \subseteq A$.
\end{proof}

@@

Notice that $\Delta_r$ is an irreducible classical root system of type $\rA_{\ell-1}$, $\rC_{\ell-1}$ or $\rD_{\ell-1}$.
The restriction of $\pi$ to the subgroup $\G(\Delta_r, R)$ decomposes into a sum of Weyl modules for $\G(\Delta_r, R)$,
 moreover, each such summand corresponds to a connected component of the weight diagram of $\pi_R$ in which all bonds marked $r$ are removed. 
Denote by $(\pi', V')$ the summand containing the highest weight vector $v^+$ of $\pi$,
 i.\,e. the summand corresponding to the top-left connected component of the diagram. Denote by $\Lambda(\pi')$ the subset of weights belonging to this component.
In all cases under consideration, $\pi'$ is isomorphic to the natural representation of $\G(\Delta_r, R)$.

Denote by $\Gamma$ the subset of weights $\lambda \in \Lambda(\pi')$ incident to the removed bonds (i.\,e. $\lambda$ such that $\lambda - \alpha_r \in \Lambda(\pi)$).
From the consideration of weight diagrams it is easy to determine the number of elements in $\Gamma$ (see~Table~\ref{table:dv-reps}).
Denote by $B$ the subset of $\E(\Delta_r, R, I)$ consisting of elements $g$ such that $(g \cdot v^+)_\lambda = 0$ for all $\lambda\in\Gamma$ and
set $A\coloneqq\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot B \cdot \EP_s(R, I) \subseteq A_{rs}.$

>>>>

As an application of parabolic factorizations, we deduce several {\it subgroup factorizations} of finite width for relative Chevalley groups.
For example, as a corollary of Bass--Kolster decomposition we obtain upper estimates of the number of factors in the presentation of a classical Chevalley group
 as a product of its subsystem subgroups $\SL(2, R, I)$, see~\cref{thm:SL2width}.
As an application of~\cref{thm:DennisVaserstein} we prove the following theorem inspired by the main result of~\cite{Nik07}.
\begin{thm}\label{thm:spin-sln-prod} Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$. \end{thm}
As another application, we obtain results on the bounded generation of the relative elementary group.
Recall from~\cite[Theorem~2]{Va86} that $\E(\Phi, R, I)$ can be generated by the set of Tits--Vaserstein generators
 $z_\alpha(s, \xi) =  x_{-\alpha}(-\xi) x_\alpha(s) x_{-\alpha}(\xi)$, where $s\in I$ and $\xi\in R$
(in fact, a smaller generating set suffices, see~\cref{prop:Stepanov-theorem}).
In section~\ref{sec:boundgen} we deduce from an earlier result of O.~Tavgen that the width $\E(\Phi, R, I)$ with respect to this generating set is
 finite for Dedekind rings of arithmetic type having a real embedding.
Moreover, as a further application of parabolic factorizations, we obtain in~\cref{thm:width} concrete estimates of such width
 for ideals of the ring $\mathbb{Z}[\sfrac{1}{p}]$ and also for ideals of stable rank $1$.

@@

\textsc{Case $\Phi=\rD_\ell$.} From the proof of \cref{item:asrUnipD} it follows that there exists $h_1 \in \EP_\ell(R, I)$ such that for $v'=h_1\cdot v$ one has $v'_2=v'_1-1\in I$.
Clearly, for $v'' = z_{-\alpha_{2}}(-v'_2, 1)\cdot v'$ one has $v''_1=1$, hence by \cref{lemma:Chevalley-Matsumoto} there exists $h_2 \in \U(\Phi^-, I)$ such that the element
$g'=h_2 \cdot z_{-\alpha_{2}}(-v'_2, 1) \cdot h_1 \cdot g$ fixes $v^+$. 
Using the Levi decomposition we can write $g'=h \cdot y \cdot x \cdot g$ for some $y\in\U(\Sigma^-_\ell, I)$, $x \in \U(\Sigma^+_\ell, I)$, $h\in\E(\Delta_\ell, R, I)$.
It is clear that $x$, $y$ are the required elements.
\end{proof}

@@

Let $G$ be a group and $A$ its subset. Denote by $L(A)$, $R(A)$ the left and the right stabilizers of $A$ and by $N(A)$ the normalizer of $A$:
\[ L(A) =  \left\{ g\in G\ \middle |\ g \cdot A = A \right\}\!,\ \ R(A) =  \left\{ g\in G\ \middle |\ A \cdot g = A \right\}\!,\ \ N(A) =  \left\{ g\in G\ \middle |\ g \cdot A \cdot g^{-1} = A \right\}.\]
It is easy to see that $L(A)$, $R(A)$ and $N(A)$ are subgroups of $G$. The following obvious lemma will be used several times in the sequel.
\begin{lemma} \label{lem:LN} The subgroup $L(A)$ is normalized by $N(A)$. Moreover, one has $R(A) \cap N(A) \subseteq L(A)$, $L(A) \cap N(A) \subseteq R(A)$. \end{lemma}
\begin{comment}
\begin{proof}
Let $h \in L(A)$, $g \in N(A)$. The first claim follows from the following chain of equalities:
\begin{equation} h^g \cdot A = g^{-1} \cdot h \cdot g \cdot A \cdot g^{-1} \cdot g = g^{-1} \cdot h \cdot A \cdot g = {A}^g = A. \end{equation} 
Now if $g \in R(A) \cap N(A)$, we obtain
$g \cdot A = g \cdot A \cdot g^{-1} \cdot g = A \cdot g = A,$ which implies the second claim.
\end{proof}
\end{comment}

@@

Denote by $A$ the product of subgroups in the right-hand side of~\eqref{eq:gauss1}.
First of all, notice that for $h \in \Hh(\Phi, R, I)$, $\beta \in \Phi$ and $\xi \in R$ one has $x_\beta(\xi)^h = x_\beta(\xi + s\xi)$ for some $s\in I$.
From this and the assumption of the proposition we obtain for $\alpha \in \Delta_1 \cup \Delta_\ell$, $\xi \in R$ that
\begin{multline} \nonumber
 A^{x_\alpha(\xi)} = (\Hh(\Phi, R, I) \cdot A(\Phi^+, I))^{x_\alpha(\xi)} \subseteq (\Hh(\Phi, R, I) \cdot \E(\Delta_i, R, I) \cdot A(\Sigma_i^+, I))^{x_\alpha(\xi)} \subseteq \\
  \subseteq \Hh(\Phi, R, I) X_{\alpha}(I) \cdot \E(\Delta_i, R, I) \cdot A(\Sigma_i^+, I) \subseteq \Hh(\Phi, R, I) \cdot \Hh(\Delta_i, R, I) A(\Delta_i^+, I) \cdot A(\Sigma_i^+, I) \subseteq A.
\end{multline}
Thus, $A$ is normalized by root subgroups $X_\alpha(R)$, $\alpha \in \Delta_1 \cup \Delta_\ell$ and therefore is normalized by $\E(\Phi, R)$.
On the other hand, for $\beta \in \Phi^+$ we have
\begin{equation} \nonumber
 X_{\beta}(I) \cdot A \subseteq \Hh(\Phi, R, I) \cdot X_{\beta}(I) A(\Phi^+, I) \subseteq A.
\end{equation}
Thus, from $\U(\Phi^+, I) \subseteq L(A)$ and~\cref{lem:LN} we obtain that $\mathcal{X} \cdot A \subseteq A$ and hence
 that $\E(\Phi, R, I) \cdot A \subseteq A$.
\end{proof}

@@

Notice that $\Delta_r$ is an irreducible classical root system of type $\rA_{\ell-1}$, $\rC_{\ell-1}$ or $\rD_{\ell-1}$.
The restriction of $\pi$ to the subgroup $\G(\Delta_r, R)$ decomposes into a sum of Weyl modules for $\G(\Delta_r, R)$,
 moreover, each such summand corresponds to a connected component of the diagram obtained from the weight diagram of $\pi_R$ by removing all bonds marked $r$. 
Denote by $(\pi', V')$ the summand containing the highest weight vector $v^+$ of $\pi$,
 i.\,e. the summand corresponding to the top-left connected component of the diagram. Denote by $\Lambda(\pi') \subseteq \Lambda(\pi)$ the subset of weights belonging to this component.
In all cases under consideration, $\pi'$ is isomorphic to the natural representation of $\G(\Delta_r, R)$.

Denote by $\Gamma$ the subset of weights $\lambda \in \Lambda(\pi')$ such that $\lambda - \alpha_r \in \Lambda(\pi)$,
 in other words, $\Gamma$ consists of weights corresponding to vertices of the weight diagram which are incident to the removed bonds.
From the consideration of weight diagrams it is easy to determine the number of elements in $\Gamma$ (see~Table~\ref{table:dv-reps}).
Denote by $B$ the subset of $\E(\Delta_r, R, I)$ consisting of elements $g$ such that $(g \cdot v^+)_\lambda = 0$ for all $\lambda\in\Gamma$ and
set $A\coloneqq\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot B \cdot \EP_s(R, I) \subseteq A_{rs}.$
